---
title: "204-asgn2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Problem setup - load data and exploratory data analysis (EDA)
We'll start by looking at the "Abalone" data. 

The data is supplied in a CSV file. Let's start by loading this into R:
```{r load data, echo=TRUE}
abalone <- read.csv("abalone.csv")
```

Now, we should split the data into two parts - a section for exploration and training, and another for testing.

```{r split abalone data, echo=TRUE}
train.idx <- sample(nrow(abalone), 0.5*nrow(abalone))
abalone.train <- abalone[train.idx, ]
abalone.test  <- abalone[-train.idx, ] ## note the negative indexing to REMOVE instances!
```

# EDA

We should start by looking at some basic descriptive information about the data:
```{r EDA.1, echo=TRUE}
 summary(abalone.train)
```


Then we are looking to plot the response (the measured number of rings):
```{r EDA.2, echo=TRUE}
hist(abalone.train$rings)
```



Next, we should examine the correlations in our data:
```{r EDA.3, echo=TRUE}
as.matrix(cor(na.omit(abalone.train[,-1])))
# cor(abalone.train)
```

```{r EDA.4, echo=TRUE}
plot(abalone.train, pch=19, col="orange")
```



